A demonstration of the \LaTeXe\ class file for \itshape{SAGE Publications}

### 2015-12-23 body/abstract.tex
geometrical feasibility

We address the issue of automatically planning discrete sequences of contacts
that characterize a multiped motion in an arbitrary cluttered environment.
To overcome the inherent combinatorial difficulty of this problem, we separate it
in two sub-problems: first, planning a guide path for the root of the robot and
then, generating relevant contacts along this path.
We propose theoretical contributions for each sub-problem, and implement them
as a complete contact planner. 
First, we propose a geometrical characterization of a guide trajectory, the ``reachability condition", which guarantees that a
guide can be mapped into the contact manifold of the robot. With this condition, asserting the``true feasibility" 
of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an off-line sampling approach,
independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
designed for generic robotic applications: a manipulatibily-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
variety of robots, including a HRP-2, Hyq, and a dexterous hand, with interactive performances.

>>>>

We address the issue of automatically planning discrete sequences of contacts
that characterize a multiped motion in an arbitrary cluttered environment.
To overcome the inherent combinatorial difficulty of this problem, we separate it
in two sub-problems: first, planning a guide path for the root of the robot and
then, generating relevant contacts along this path.
We propose theoretical contributions for each sub-problem, and implement them
as a complete contact planner. 
First, we propose a geometrical characterization of a guide trajectory, the ``reachability condition", which guarantees that a
guide can be mapped into the contact manifold of the robot. With this condition, asserting the``true feasibility" 
of a guide path boils down to performing collision tests, thus avoiding the expensive explicit computation of balanced contact configurations, as opposed to previous approaches.
Then, we propose a sequential algorithm to extend a truly feasible path into a sequence of contacts. For each limb, an off-line sampling approach,
independent from the environment, is run once and for all, and used at run time to provide efficiently candidate configurations for contact.
The best candidate is chosen according to user-defined heuristics. We propose two such heuristics, used in all our experiments, and 
designed for generic robotic applications: a manipulatibily-based heuristic, EFORT, and a robustness criterion regarding the balance of the robot.
These contributions are implemented within the open-source software Hpp. Our framework allows us to generate complex multi-contact sequences for a large
variety of robots, including HRP-2, Hyq, and a dexterous hand, with interactive performances.

### 2015-12-23 body/contact.tex
typos
geometrical feasibility

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
As an input of this stage, we consider a geometrically feasible guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$. We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.

@@

In what follows a \textit{valid} contact configuration is understood as being collision free, satisfying joint limits, and allowing to maintain the static equilibrium
of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
is described in Section TODO.

\subsection{Definition of a contact generator}
\label{sec:single_contact}
We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$. TODO FIGURE
From true feasibility we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
The issue is then to evaluate locally at least one element of $P$ at $\mathbf{q}^0$.

@@

We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
between an effector and a surface of the environment is less than a small value $\epsilon$.
We then apply the following steps:
\begin{enumerate}
\item Generate offline $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree stata structure;
\item Figure~\ref{fig:contact_gen}--b and c--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item Figure~\ref{fig:contact_gen}--d and e--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
\item If a valid solution is found, stop. Otherwise remove the element from $S$ and go back step 5.
\end{enumerate}

This approach has two main advantages.
First, this allows us to select a large number of candidates in a single proximity request.
Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
a locally-optimal candidate.
Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
Regarding completeness, it is immediate to verify that as $N$ and $\epsilon$ grow, the completeness of the method is restored.
$N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.

@@

These three rules provide a deterministic manner to generate contacts along the discretized guide path.
TODO FIGURE. For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}, thanks to 
the heuristics that we use for contact selection.

>>>>

% !TEX root =  ../main.tex
\section{From a guide path to a discrete sequence of contact configurations ($\mathcal{P}_2$)}
\label{sec:contact}
As an input of this stage, we consider a \textit{geometrically feasible} guide path $\mathbf{q}^0(t) : [0,1] \longrightarrow SE(3)$ for the root of the robot $R$. We now consider the second problem of computing a discrete sequence of contact configurations $\mathbf{Q}^{\overline{0}}$ for the limbs of the robot along $\mathbf{q}^0(t)$.

@@

In what follows a \textit{valid} contact configuration is understood as being collision free, satisfying joint limits, and allowing to maintain the static equilibrium
of the robot. The first two tests are trivial; our criterion to assert efficiently the static equilibrium of the system
is described in Section~\ref{sec:heuristics}.

\subsection{Definition of a contact generator}
\label{sec:single_contact}
We consider a limb $R^k$  and a root configuration $\mathbf{q}^0$.
From \textit{geometrical feasibility} we know that there exists an infinite list of mappings $P = [\pi_0,..,\pi_i, ...]$ that verify \eqref{eq:pi} for $\mathbf{q}^0$.
The issue is then to evaluate locally at least one element of $P$ at $\mathbf{q}^0$.

@@

We first define $C_{Contact}^{\epsilon} \supset C_{Contact}$ as the set of configurations almost in contact. This means that the minimal distance 
between an effector and a surface of the environment is less than a small value $\epsilon$.
We then apply the following steps:
\begin{enumerate}
\item Generate off-line $N$ valid sample limb configurations $\mathbf{q}^k_i,  0 \leq i < N$;
\item Using the end-effector positions $\mathbf{p}(\mathbf{q}^k_i)$ as indices, store each sample in an octree stata structure;
\item Figure~\ref{fig:contact_gen}--b and c--: At runtime, when contact creation is required, retrieve from the octree the list of samples $S \subset C_{Contact}^{\epsilon}$ close to contact;
\item According to a user-defined heuristic $h$, sort the elements of $S$;
\item Figure~\ref{fig:contact_gen}--d and e--: Select the first configuration of $S$ (or fail if $S$ is empty). Project the configuration onto contact with inverse kinematics;
\item If a valid solution is found, stop. Otherwise remove the element from $S$ and go back step 5.
\end{enumerate}

This approach has two main advantages.
First, this allows us to select a large number of candidates in a single proximity request.
Having several candidates is interesting, because it allows to compare them using a user-selected heuristic $h$, thus obtaining
a locally-optimal candidate.
Furthermore, the fact that the candidates are already close to contact increases the odds that the inverse kinematics will converge to a valid solution in a small number of iterations.
Regarding completeness, it is immediate to verify that as $N$ grows, the completeness of the method is restored.
$N$ is a parameter that allows us to specify the trade-off between completeness and efficiency.
The reader is invited to refer to our previous work for an extensive discussion on the optimal value of $N$ \citep{Tonneau2014}.

@@

These three rules provide a deterministic manner to generate contacts along the discretized guide path.
For a given truly feasible path, our current implementation does not guarantee that the planner will succeed in generating a configuration in static equilibrium, because
of the deterministic approach used to break the combinatorial.
However, in practice the planner is successful in the large majority of cases, as discussed in Section~\ref{sec:perf}, thanks to 
the heuristics that we use for contact selection.

### 2015-12-23 body/efort.tex
geometrical feasibility

We assume that $J$ is full rank (this is reasonable since singular configurations are discarded by our algorithm).
From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
\begin{equation} \label{jaceq}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
\end{equation}

@@

Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
\begin{equation*} \label{ballforce}
||\mathbf{\tau}||^2 \leq 1
\end{equation*}
is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right), defined by:
\begin{equation} \label{ellipsoidforce}
\mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
\end{equation}

>>>>

We assume that $\mathbf{J}$ is full rank (we are not interested in singular configurations that we discard).
From \eqref{jac} we can thus obtain the following equality (Appendix~\ref{app:manipulability}):
\begin{equation} \label{jaceq}
\mathbf{\dot{p}}^T(\mathbf{J}\mathbf{J}^T)^{-1}\mathbf{\dot{p}} = \dot{\mathbf{q}}^T \dot{\mathbf{q}}
\end{equation}

@@

Exploiting \eqref{jac}, we can easily see that the set of achievable forces in $\mathbb{R}^m$ subject to the constraint:
\begin{equation*} \label{ballforce}
||\mathbf{\tau}||^2 \leq 1
\end{equation*}
is the so-called force ellipsoid (Figure~\ref{sec:efort_ellipsoid} - right):
\begin{equation} \label{ellipsoidforce}
\mathbf{f}^T (\mathbf{J}\mathbf{J}^T) \mathbf{f} \leq 1
\end{equation}

### 2015-12-23 body/equilibrium.tex
typos
geometrical feasibility

\subsubsection{Conditions for static balance.}
We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

@@

Any force belonging to the linearized cone
can thus be expressed as a convex combination of its four generating rays.
%~ \deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}

>>>>

\subsubsection{Conditions for static balance.}
We first define the variables of the problem, for a robot $R$ and $e$ contact points, expressed in world coordinates:
\begin{itemize}
\item $\mathbf{c} \in \mathbb{R}^3$ is the robot center of mass (COM);
\item $\mathbf{L}  \in \mathbb{R}^3 $ is the angular momentum at the COM;
\item $\mathbf{m} \in \mathbb{R}$ is the robot mass;
\item $\mathbf{g} = [0,0,-9.81]^T$ is the gravity acceleration;
\item $\mu$ is the friction coefficient;
\item for the i-th contact point $1 \leq i \leq e$:
	\begin{itemize}
	\item $\mathbf{p_i}$ is the contact position;
	\item $\mathbf{f_i}$ is the force applied at $\mathbf{p_i}$;
	\item $\mathbf{n}_i,\mathbf{t}_{i1},\mathbf{t}_{i2}$ form a local Cartesian coordinate system centered at $\mathbf{p_i}$. $\mathbf{n}_i$ is aligned
	with the contact surface normal, and the $\mathbf{t}_i$s are tangent vectors.
	\end{itemize}
\end{itemize}

@@

Any force belonging to the linearized cone
can thus be expressed as a positive combination of its four generating rays.
%~ \deleted{Therefore we can express the nonslipping constraint on $\mathbf{f}_i$ as:}

### 2015-12-23 body/intro.tex
geometrical feasibility

The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper. The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: the planning of a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).  A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments. Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

@@

\subsection{Paper contribution and organization}
This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
we believe that the separation between the generation of the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated:
\begin{itemize}
\item The guide trajectory must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces). We call this property \textit{true feasibility}. This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}.
\item There are infinite combinations of possible contact sequences for a given root trajectory. The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions.
\end{itemize}

By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
For each problem, we introduce an original formulation of the related theoretical issues.
Then, we propose a concrete implementation of the solution, where the constraints of the problem are relaxed in favor of efficiency.
The relaxation is justified empirically, with statistics presented for 2 robots and 4 scenarios. The complete framework is implemented within
the open-source platform HPP.

@@

To address \Pa we rigorously formulate a geometric condition that approximates the true feasibility of a path, the reachability condition. We claim that this 
condition can be verified in a low-dimensional space, which we call $C_{reach}$. This formulation can make the planning of a guide path more efficient computationally, while providing similar guarantees to planning directly in the configuration space.
In practice, for each considered robot, we compute an approximation of $C_{reach}$. From this approximation,
an efficient trade-off between a necessary and a sufficient condition for true feasibility is obtained. This trade-off allows the implementation of a low-dimensional, Reachability-Based RRT planner (RB-RRT), able to find truly-feasible paths for the root of the robot in a timespan ranging from a few milliseconds to a few seconds, depending on the scenario.

To address \Pb, we consider a truly-feasible path as an input, and generate new contact configurations along it using a sampling approach.
The true feasibility of the path provides guarantees that the contacts can be created, and reduce the dimensionality of the problem (at each step the root position is fixed). The theoretical completeness of the approach is partially traded for efficiency: instead of sampling on-line
limb configurations and projecting them onto contact, we precompute off-line a large database of limb configurations, stored in a spatial structure optimized
for proximity queries. At runtime, this framework allows to obtain rapidly a set of candidate configurations for contact.
These candidates are sorted using a criterion to evaluate the static equilibrium of the configuration, as well as a set of user-defined heuristics.
In all our experiments we used two such heuristics: a measurement of robustness of a configuration to errors in the contact forces, and a manipulability-based heuristic,
the Extended FORce Transmission ratio (EFORT).

The contribution of the paper is thus twofold. We propose a theoretical characterization of today's most efficient practical approach to sampled-based planning of acyclic contacts. Based on this characterization, we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications.

The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep(tonneauisrr15).
The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with a discussion on the implementation of this theory to real-world robots and problems.
In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the HPP implementation
of our contact generation algorithm. Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}. Appendix~\ref{app:rom} describes the generation process
for the robot.

In Section~\ref{overview}, we present the general organization of our method. Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb. In Section~\ref{sec:heuristics}, we present two heuristics used for the selection of a contact configuration. Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
that complete those presented in the conference paper.

>>>>

The issue of planning acyclic contacts was first completely described by \citeauthor{conf/iser/BretlRLKA04} in their seminal paper. The issue requires the simultaneous handling of two problems, $\mathcal{P}_1$: Planning a relevant guide trajectory for the root of the robot in $SE(3)$; and $\mathcal{P}_2$: Planning a discrete sequence of acyclic contact configurations along the trajectory (A third nontrivial problem, $\mathcal{P}_3$, not addressed in this work, then consists in interpolating a complete motion between two postures of the contact sequence).  A key issue is to avoid combinatorial explosion when considering at the same time the possible contact configurations and the potential trajectories. This seminal paper proposes a first effective algorithm, able to handle simple situations (such as climbing scenarios), but not applicable to arbitrary environments. Following it, seve\-ral papers have applied this approach in particular situations, typically limiting the combinatorial by imposing a fixed set of possible contacts \citep{Hauser06usingmotion, stilman2010}.

@@

\subsection{Rationale}
This paper presents a pragmatic approach to break the complexity of multi-contact planning. With that objective in mind,
we believe that the separation between the generation of the guide trajectory and the contact sequence is the most promising direction \citep{DBLP:conf/iser/EscandeKMG08}.
However, this direction raises two theoretical questions that remain to be solved, or even to be properly formulated. \\

\noindent \textbf{Regarding $\mathcal{P}_1$}, the guide trajectory must guarantee the existence of a contact sequence to actuate it (This property is related to the controllability of the root actuated by the contact forces). We call this property \textit{true feasibility}. This property has not been studied yet; the only way to validate a waypoint in the trajectory is to explicitly compute the contact locations and forces, which is computationally not reasonable \citep{Bouyarmane2009}, unless the scenario is limited to cyclic, quasi-flat cases \citep{zucker2010optimization}. \\

\noindent \textbf{Regarding $\mathcal{P}_2$}, there are infinite combinations of possible contact sequences for a given root trajectory. The selection of one particular contact sequence with interesting properties (minimum number of contact changes, robustness, efficiency or naturalness) has been studied for cyclic cases \citep{Hauser06usingmotion}, but has not been efficiently applied to cluttered environments (\citeauthor{bouyarmane:lirmm-00777727, DBLP:conf/iser/EscandeKMG08} mostly randomly picked one contact sequence, leading to possibly very tedious transitions).  \\

%~ In this paper, we address both questions.
%~ For each problem, we introduce an original formulation of the related theoretical issues.
%~ Then, we propose a concrete implementation of a solution, where the constraints of the problem are relaxed in favor of efficiency.

The key idea of the paper is the theoritical and practical characterization of the space of \textit{truly feasible} configurations to address both issues.
We first exhibit a low dimensional over-approximation of this space. This approximation is defined geometrically. 
We make the strong assumption that these \textit{geometrically feasible} paths can be efficiently transformed into \textit{truly feasible} paths
in cluttered environments. 
Under this assumption we can take advantage of the decoupling approach.

We first address \Pa by planning \textit{geometrically feasible} paths. We achieve this efficiently thanks to the \textit{reachability condition} that we introduce, that 
does not require explicit contact computation.

Then we address \Pb by extending the path into a \textit{truly feasible} sequence of contact configurations.
This second step requires computing contact configurations verifying a condition for \textit{static equilibrium feasibility}, that requires
explicit contact computation.

This sequential approach is the key to the efficiency of our method, though it can
can result in failures because our assumption is not always true. However, we demonstrate empirically the validity of the approach: the high success rate of our method, combined with the low computation times we obtain, allows us to plan (and re-plan upon failure) multi contact sequences at interactive rates.

\subsection{Paper contribution and organization}

By tackling these issues, our method proposes contributions to both problems \Pa and \Pb.
\begin{itemize}
\item First, we propose a theoretical characterization of an efficient approach to sampling-based planning of acyclic contacts;
\item Then we propose a very efficient and general implementation of an acyclic contact planner, the first one compatible with interactive applications;
\item Finally we propose three heuristics for contact generation. They bias the planner towards configurations that are robust regarding static equilibrium, or 
efficient regarding the task being performed.
\end{itemize}

@@

In Section~\ref{overview}, we present the general organization of our method. Section~\ref{rbprm} and Section~\ref{sec:contact} present respectively our answer to \Pa and \Pb. In Section~\ref{sec:heuristics}, we present two heuristics used for the selection of a contact configuration. Finally, we propose a complete experimental validation of the planner with three very different kinematic chains (the HRP-2 and HyQ robots, and a three-finger manipulator) in various scenarios,
that complete those presented in the conference paper.
\subsection*{Comparison with our previous work.}
The present paper is an extension of a conference paper to appear in the proceedings of the ISRR'15 conference~\citep{tonneauisrr15}.
The conference paper focuses on the theoretical formulation of the problem, and presents results obtained with virtual avatars.
This extension completes this work with a discussion on the implementation of this theory to real-world robots and problems.
In particular, we introduce a robust balance criterion, designed to ensure the equilibrium of the robot despite bounded error in the contact forces. We also provide details on the HPP implementation
of our contact generation algorithm. Furthermore, our experiments are now applied to real-robot models, namely HRP-2 and HyQ\adnote{cite Semini}. Appendix~\ref{app:rom} describes the generation process
for the robot.

### 2015-12-23 body/overview.tex
geometrical feasibility

Figure~\ref{fig:framework} illustrates our workflow.
\Pa and \Pb are addressed in a sequential fashion.
The guarantees provided by the truly-feasible guide path allow
us to avoid the combinatorial by generating the contacts in a straightforward manner.

%
\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
		\put (1,1) {a} 
		\put (22,1) {b} 
		\put (42,1) {c} 
		\put (62,1) {d} 
		\put (83,1) {e} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right leg of HRP-2. a: Selection of reachable obstacles. b: Entries of the limb samples database (with $N = 4$). c: With a proximity query on the octree database, configurations too far from obstacles are eliminated. d: The best candidate according to a user-defined heuristic $h$ is chosen. e: The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure*}
\subsection{Computation of a guide path --- \Pa}
We first consider the problem of planning a relevant guide path. The objective is to compute a path of root placements that will allow contact creation. A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be truly feasible, i.e. must lead to a valid sequence of contacts.

More precisely, a root placement is truly feasible if and only if there exists at least one joint configuration such that the robot satisfies three conditions:
\begin{description}
\item[C1] is not in collision with the environment, 
\item[C2] has at least one end-effector in contact with the environment, 
\item[C3] is in static equilibrium. 
\end{description}
The conditions C1 and C2 are purely geometrical and can be efficiently approximated even without computing the contact points. Instead C3 is difficult to approximate and it requires the computation of the contact points, and the resolution of a linear program to compute the contact forces~\citep{Prete2016}. Moreover our intuition suggests us that, for cluttered environment, C3 is not critical, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3. For this reason we neglect C3 during the resolution of \Pa, and we worry about it only when solving \Pb.

An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the Reachable workspace of the robot (represented for the right arm in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision. We define formally $C_{reach}$, the set of root placements satisfying C1 and C2, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa). Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
%
Details are presented in Section~\ref{rbprm}.

@@

The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--\Pb). In this
step we assume that the input guide path is truly feasible.
With a fixed root position, the dimensionality of the 
contact generation problem is reduced to the number of degrees of freedom of the considered limb. In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, which is independent from the environment (Figure~\ref{fig:contact_gen}).

>>>>

Figure~\ref{fig:framework} illustrates our workflow.
\Pa and \Pb are addressed in a sequential fashion.
The guarantees provided by the \textit{geometrically feasible} guide path allow
us to avoid the combinatorial by generating the contacts in a straightforward manner.

%
\begin{figure*}
  \centering
  \begin{overpic}[width=0.8\linewidth]{figures/contact_gen}
		\put (1,1) {a} 
		\put (22,1) {b} 
		\put (42,1) {c} 
		\put (62,1) {d} 
		\put (83,1) {e} 
	\end{overpic}
  \caption{Generation of a contact configuration for the right leg of HRP-2. a: Selection of reachable obstacles. b: Entries of the limb samples database (with $N = 4$). c: With a proximity query on the octree database, configurations too far from obstacles are eliminated. d: The best candidate according to a user-defined heuristic $h$ is chosen. e: The final contact is achieved using inverse kinematics.}
  \label{fig:contact_gen}
\end{figure*}
\subsection{Computation of a guide path --- \Pa}
We first consider the problem of planning a relevant guide path. The objective is to compute a path of root placements that will allow contact creation. A desired requirement is to preserve the completeness of the planner: it should be able to explore any possible guide trajectory, but at the same time, any computed guide trajectory must be \textit{truly feasible}, i.e. must lead to a valid sequence of contacts.

More precisely, a root placement is \textit{truly feasible} if and only if there exists at least one joint configuration such that the robot satisfies three conditions:
\begin{description}
\item[C1] is not in collision with the environment, 
\item[C2] has at least one end-effector in contact with the environment, 
\item[C3] is in static equilibrium. 
\end{description}
C1 and C2 define \textit{geometrical feasibility}, and can be efficiently approximated even without computing the contact points. C3 defines \textit{equilibrium feasibility}. C3 is difficult to approximate and requires the computation of the contact points, as well as the resolution of a linear program to compute the contact forces~\citep{Prete2016}. 
However our intuition suggests us that, for cluttered environments, verifying C3 is not critical for $\mathcal{P}_1$, meaning that, most of the times, root placements satisfying C1 and C2 would satisfy also C3. For this reason we neglect C3 during the resolution of \Pa, and we worry about it only when solving \Pb.

An intuitive description of root placements satisfying C1 and C2 is ``close, but not too close'': close, because a contact surface must be partially included in the reachable workspace of the robot (represented for the right arm in Figure~\ref{fig:contact_gen}--a); not too close, because the robot must avoid collision. We define formally $C_{reach}$, the set of root placements satisfying C1 and C2, in which we compute a guide trajectory with a sampling-based planner, the Reachability-Based RRT---RB-RRT--- (Figure~\ref{fig:framework}---\Pa). Planning in $C_{reach}$ boils down to planning in SE(3), which has an acceptable practical complexity.
%
Details are presented in Section~\ref{rbprm}.

@@

The second stage extends the guide path into a sequence of contact configurations (Figure~\ref{fig:framework}--\Pb). In this
step the input guide path is \textit{geometrically feasible}.
With a fixed root position, the dimensionality of the 
contact generation problem is reduced to the number of degrees of freedom of the considered limb. In other words, we consider each limb as a manipulator attached to the root, and select the most relevant contact from a database of precomputed configurations, which is independent from the environment (Figure~\ref{fig:contact_gen}).
